We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-21 paper/paper.tex
edit

Later, \cite{2007MNRAS.374.1125M} used simulations to confirm the low
chances ($<0.5\%$) found in \cite{2005A&A...431..517K}.
This result was challenged by \cite{2009MNRAS.399..550L} who continue
to report high chances to find a planar configuration;
two more recent numerical experiment with high resolution simulations
(\cite{2013MNRAS.429..725S} and \cite{2016MNRAS.457.1931S}) report
contradicting results (low and high chances to find the observational
result, respectively) but precise figures are not quoted in either of
these three reports. \citep{2015ApJ...815...19P,2016MNRAS.457.1931S}.
Meanwhile, \cite{2013MNRAS.429.1502W, 2014ApJ...789L..24P} reported
on new tests using high resolution simulations and cosmological
volumes arriving at a chance of $6\%$ and $0.77\%$, respectively to
have a satellite distribution as triaxial as the observations.

In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
total poulation of $27$ satellites \citep{2013ApJ...766..120C,
2013Natur.493...62I}, although its brightest satellites
do not show any special feature.
All the studies agree on the point that the satellite distribution of
the 15 to 27 brightest satellites are consistent with 
spherically symmetric distribution and easy to reproduce in LCDM
simulations \citep{2006AJ....131.1405K,
  2007MNRAS.374.1125M, 2013ApJ...766..120C}.

\begin{table*}
\centering
\begin{tabular}{|p{3.0cm}|p{6cm}| p{4.5cm}| c|}\hline
Reference & Target Measurement & Parent Sample & Probability ($\%$)\\\hline
\text{\cite{2006AJ....131.1405K}} & RMS height in M31's 15 brightest
satellites & Monte Carlo satellite distributions with a power law
radial distribution & $87-99$\\
\text{\cite{2007MNRAS.374.1125M}} & Triaxiality and RMS height in M31's
16 brightest satellites & Monte Carlo from an spherical power law radial
distribution & $17$\\
\text{\cite{2013ApJ...766..120C}}& RMS height in M31's 27 brightest
satellites & Monte Carlo randomized satellite distribution & High\\
This Work & Triaxiality and RMS height in M31's 11-15 brightest
satellites & 24 halos from selected pairs in a cosmological N-body Dark Matter only simulation ($\sim
10^{6}$\Msun particle mass resolution)& 56 \\
This Work & Triaxiality and RMS height in M31's 11-15 brightest
satellites & 20 halos from selected pairs in a cosmological N-body hydro simulation ($\sim
10^{6}$\Msun particle mass resolution)& 27 \\
This Work & Triaxiality and RMS height in M31's 11-15 brightest
satellites & 12 high resolution DM only N-body simulations ($\sim
10^{5}$\Msun particle mass resolution) of halo pairs & 37 \\
\hline
\end{tabular}
\caption{Probability results derived in previous publications for the
  triaxiality and/or root mean squared (RMS) height of M31 satellites.
\label{table:M31}}
\end{table*}

@@

The aim of our work is to overcome two difficulties in the study of
atipicallity in satellite distributions.
First, atypicallity is usually computed from distributions comming from
different simulated pair samples, making it inconvenient to compare
results from different studies. 
Second, halo pairs similar to the LG are scarse in cosmological
volumes, making it hard to perform direct comparisons against simulated
systems that fully resemble the LG.

We surmount the first difficulty by setting as a point of reference a
spherical satellite distribution and not LCDM simulations.
%comentario: esta explicación no me parece muy clara
This spherical distribution is built from the data itself
(observational or simulated) by randomizing its angular position
around the central galaxy and keeping its radial distance fixed
\citep{2017AN....338..854P}. 
We characterize each satellite distribution with its inertia
tensor and summarize the results with the scalars describing the width
of the distribution and its axis ratios.

>>>>

Later, \cite{2007MNRAS.374.1125M} used simulations to confirm the low
chances ($<0.5\%$) found in \cite{2005A&A...431..517K}.
This result was challenged by \cite{2009MNRAS.399..550L} who continue
to report high chances to find a planar configuration;
two more recent numerical experiment with high resolution simulations
(\cite{2013MNRAS.429..725S} and \cite{2016MNRAS.457.1931S}) report
contradicting results (low and high chances to find the observational
result, respectively) but precise figures are not quoted in either of
these three reports. \citep{2015ApJ...815...19P,2016MNRAS.457.1931S}.
Meanwhile, \cite{2013MNRAS.429.1502W} and \cite{2014ApJ...789L..24P} reported
on new tests using high resolution simulations and cosmological
volumes arriving at a chance of $6\%$ and $0.77\%$, respectively, to
have a satellite distribution as triaxial as the observations.

In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
total poulation of $27$ satellites \citep{2013ApJ...766..120C,
2013Natur.493...62I}, although its brightest satellites
do not show any special feature.
All published studies agree on the point that the global satellite distribution of
the 15 to 27 brightest satellites are consistent with 
spherically symmetric distribution and easy to reproduce in LCDM
simulations \citep{2006AJ....131.1405K,
  2007MNRAS.374.1125M, 2013ApJ...766..120C}.    
Table \ref{table:M31} and Table \ref{table:MW} summarize the story we
have just told, including the results from this paper using the
methodolgy we describe in the upcoming sections.

\begin{table*}
\centering
\begin{tabular}{|p{3.0cm}|p{6cm}| p{4.5cm}| c|}\hline
Reference & Target Measurement & Parent Sample & Probability ($\%$)\\\hline
\text{\cite{2006AJ....131.1405K}} & RMS height in 15 brightest
satellites & Monte Carlo satellite distributions with a power law
radial distribution & $87-99$\\
\text{\cite{2007MNRAS.374.1125M}} & Triaxiality and RMS height in 16 brightest satellites & Monte Carlo from an spherical power law radial
distribution & $17$\\
\text{\cite{2013ApJ...766..120C}}& RMS height in M31's brightest
satellites & Monte Carlo randomized satellite distribution & High\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 24 halos from selected pairs in a cosmological N-body Dark Matter only simulation ($\sim
10^{6}$\Msun particle mass resolution)& 56 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 20 halos from selected pairs in a cosmological N-body hydro simulation ($\sim
10^{6}$\Msun particle mass resolution)& 27 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 12 high resolution DM only N-body simulations ($\sim
10^{5}$\Msun particle mass resolution) of halo pairs & 37 \\
\hline
\end{tabular}
\caption{Probability results derived in previous publications for the
  triaxiality and/or root mean squared (RMS) height of M31 satellites.
\label{table:M31}}
\end{table*}

@@

Trying to reconcile and understand the seemly conflicting or inconclusive
results on the MW has its origin on the frequentist fashion generally used to
compute probabilities.
Usually, the process starts by building a high level parent sample in the
simulation and then counting how many elements has a subset
meeting some criteria. 
The first inconvenient with this approach is that the probability
estimate is made against whatever turns out to be typical in each
simulation. 
Comparing results across simulations would require first
characterizing the simulations at the high level parent samples.
The second inconvenient is that it is hard to find systems that fully resemble
the LG, this means that the high level parent sample has a small size
making it hard to find elements that fit some criteria.

In this paper, we surmount the first difficulty by setting as a point of reference a
spherical satellite distribution and not LCDM simulations.

This spherical distribution is built from the data itself
(observational or simulated) by randomizing its angular position
around the central galaxy and keeping its radial distance fixed
\citep{2017AN....338..854P}. 
We characterize each satellite distribution with its inertia
tensor and summarize the results with the scalars describing the width
of the distribution and its axis ratios.
